---
output: github_document
always_allow_html: true
---

<!-- README.md is generated from README.Rmd. Please edit that file -->
```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# epidscan

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

**A Modern R Interface for Spatio-Temporal Scan Statistics.**

`epidscan` brings modern data science workflows to epidemiological cluster detection. Currently powering [SaTScan™](https://www.satscan.org/), it is designed to be a flexible, type-safe, and stateless interface for spatial and space-time analysis.

## Design Principles

-   **Stateless**: No global side-effects.
-   **Type-Safe**: `ss_tbl` classes ensure your data is analysis-ready before you run a single line of code.
-   **Spatial-Native**: Built-in support for `sf` simple features.

## Installation

```r
# Install from GitHub
pak::pak("Songyosr/epidscan")
```

**Requirements:** [SaTScan™](https://www.satscan.org/) must be installed on your system.

## Quick Start

Here is a complete space-time analysis of the New Mexico lung cancer dataset (included in the package).

```{r example}
library(epidscan)
library(dplyr)
library(broom)

# 1. Prepare Data
# Load included datasets
data("NMlung_cas", "NMlung_pop", "NMlung_geo")

# Create validated inputs
cas <- ss_cas(NMlung_cas, loc_id = "county", cases = "cases", time = "date", time_precision = "month", covars = c("age_group", "sex"))
pop <- ss_pop(NMlung_pop, loc_id = "county", population = "population", time = "year", time_precision = "year", covars = c("age_group", "sex"))
geo <- ss_geo(NMlung_geo, loc_id = "county", coord1 = "x_km", coord2 = "y_km", coord_type = "cartesian")

# 2. Run Analysis
# Set path to SaTScan (adjust for your system)
# Note: This path is specific to macOS; adjust for Windows/Linux
# set_satscan_path("/Applications/SaTScan.app/Contents/app/satscan")
set_satscan_path("/Applications/SaTScan.app/Contents/app/satscan")

# Configure with prm_options (Object-Oriented Parameter Management)
# This creates a reproducible parameter object that can be inspected and reused.
my_prm <- prm_options(
  AnalysisType = 3, # Retrospective Space-Time
  ModelType = 0, # Poisson
  MonteCarloReps = 999
)

# Run the analysis
result <- satscanr(
  cas, pop, geo,
  prm = my_prm,
  verbose = FALSE
)

# 3. Explore Results
# Tidy summary of detected clusters
summary(result)

# Analysis-level metrics
tidy(result)

# Visualization
# map_clusters(result, simple = TRUE) # Interactive map (view locally)
```

For a detailed walkthrough, implementation details, and advanced configuration, see the [Getting Started vignette](vignettes/getting-started.Rmd).

## Available Input Types

| Function | File Type | Purpose | status |
|----------|-----------|---------|:---:|
| `ss_cas()` | `.cas` | Case counts | ✓ |
| `ss_pop()` | `.pop` | Population denominators | ✓ |
| `ss_geo()` | `.geo` | Spatial coordinates | ✓ |
| `ss_ctl()` | `.ctl` | Controls (Bernoulli) | ✓ |
| `ss_grd()` | `.grd` | Custom scan centers | ✓ |
| `ss_nwk()` | `.nwk` | Network edges | Under Dev |
| `ss_nbr()` | `.nbr` | Neighbor rankings | Under Dev |
| `ss_met()` | `.met` | Meta-locations | Under Dev |

## Documentation

- [Getting Started](doc/getting-started.html)
- [Parameter Reference](doc/satscan-parameters.html)
-   [Report Issues](https://github.com/Songyosr/epidscan/issues)

## Citation

Please cite the original SaTScan software and methodology:

> Kulldorff M. (1997). A spatial scan statistic. Communications in Statistics: Theory and Methods, 26:1481-1496.

> Kulldorff M. and Information Management Services Inc. (2018). SaTScan™ User Guide. www.satscan.org.

To cite this R package:

```r
citation("epidscan")
```

## License

MIT © Songyos Rajborirug
