---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->
```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# epidscan

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

A modern R interface for [SaTScan™](https://www.satscan.org/) spatial scan statistics. Provides type-safe data structures, parameter management, and native spatial data support for epidemiological cluster detection.

## Design Principles

`epidscan` is built around a few core ideas:

**Stateless by design:** No hidden global state. Every function call is independent and reproducible. Input validation happens upfront rather than at runtime.

**Type-safe inputs:** The `ss_tbl` system validates data structure before analysis runs. Location IDs, coordinates, and time formats are checked at construction time.

**Spatial-first:** Native support for `sf` objects. Coordinate system detection and ordering handled automatically.

**Parameter transparency:** Three-level parameter hierarchy (data defaults → function overrides → template files) makes configuration explicit and traceable.

## Installation
```r
# Install from GitHub
pak::pak("Songyosr/epidscan")
```

**Requirements:** [SaTScan™](https://www.satscan.org/) must be installed on your system.

## Example

Using the New Mexico brain cancer dataset (distributed with SaTScan):
Using the New Mexico Lung Cancer dataset (included in `epidscan`):
```{r example}
library(epidscan)
library(dplyr)
library(broom)

# Configure SaTScan path (one time)
ss_path <- "/Applications/SaTScan.app/Contents/app/satscan" # Depend on yours location
set_satscan_path(ss_path)

# Load the New Mexico Lung Cancer data
data(NMlung_cas)
data(NMlung_pop)
data(NMlung_geo)

# Check data
head(NMlung_cas)
head(NMlung_pop)
head(NMlung_geo)

# Prepare inputs
# Case File: Monthly data with Age/Sex covariates
cas <- ss_cas(
  NMlung_cas,
  loc_id = "county",
  cases = "cases",
  time = "date",
  time_precision = "month",
  covars = c("age_group", "sex")
)

# Population File: Yearly census data
pop <- ss_pop(
  NMlung_pop,
  loc_id = "county",
  time = "year",
  population = "population",
  time_precision = "year",
  covars = c("age_group", "sex")
)

# Coordinates File: Cartesian (km)
geo <- ss_geo(
  NMlung_geo,
  loc_id = "county",
  coord1 = "x_km",
  coord2 = "y_km",
  coord_type = "cartesian"
)

# Run analysis
# Create parameter object (Auto-detects SaTScan version)
my_prm <- prm_options(
  AnalysisType = 3, # RetroSpective Space-time
  ModelType = 0, # Poisson
  TimeAggregationUnits = 3, # Month
  MonteCarloReps = 999
)

# Preview configuration
print(my_prm)

result <- satscanr(
  cas = cas,
  geo = geo,
  pop = pop,
  prm = my_prm,
  output_dir = "example_output"
)

# Examine results
summary(result)

library(broom)
tidy(result)

glance(result)
```

```{r map, eval=FALSE}
# simple = TRUE for Cartesian/blank canvas mapping
map_clusters(result, simple = TRUE)
```

## Core Workflow

### 1. Create validated inputs

The `ss_*()` functions construct type-safe input objects:
```{r eval=FALSE}
# Case file
cas <- ss_cas(
  data,
  loc_id = "location_id",
  cases = "case_count",
  time = "onset_date",
  time_precision = "day" # "day", "month", "year", or "generic"
)

# Coordinates file (works with sf objects or data frames)
geo <- ss_geo(
  shapefile_sf,
  loc_id = "location_id"
  # Centroids extracted automatically
  # CRS detected and coordinates ordered correctly for SaTScan
)

# Population file
pop <- ss_pop(
  population_data,
  loc_id = "location_id",
  time = "census_year",
  population = "pop_count",
  time_precision = "year"
)
```

### 2. Configure & Run
```{r eval=FALSE}
# New: Object-Oriented Configuration
opts <- prm_options(
  AnalysisType = 3, # Space-time
  MonteCarloReps = 999,
  MaxSpatialSizeInPopulation = 50
)

result <- satscanr(
  cas,
  geo = geo,
  pop = pop,
  prm = opts
)
```

The `satscanr()` function:
- Writes SaTScan input files (`.cas`, `.geo`, `.pop`)
- Constructs parameter file using three-level hierarchy
- Executes SaTScan
- Parses results into structured R objects

### 3. Work with results

Results implement broom methods for tidy analysis:
```{r eval=FALSE}
# Cluster-level details
tidy(result)

# Analysis-level summary
glance(result)

# Visualize clusters
# The NM dataset uses arbitrary X/Y coordinates (named lat/long).
# We set simple = TRUE to plot them on a blank canvas (Simple CRS).
map_clusters(result, simple = TRUE)
```

## Key Features

### The `ss_tbl` System

Input validation at construction time:
```{r eval=FALSE}
cas <- ss_cas(data, loc_id = "id", cases = "n", time = "date")
class(cas)
#> [1] "ss_tbl"      "data.frame"

print(cas)
#> <ss_tbl: cas>
#> Roles:
#>   - loc_id -> id
#>   - cases -> n
#>   - time -> date
#> Spec:
#>   - time_precision: day
#> Data: 1,234 rows x 3 cols
```

Features:
- Zero-case rows automatically excluded
- Time formatting deferred to write-time based on precision
- Column role mappings explicit and inspectable

### Spatial Data Support

Native `sf` integration:
```{r eval=FALSE}
library(sf)

# From sf object
geo <- ss_geo(my_shapefile, loc_id = "GEOID")
# → Extracts centroids
# → Detects coordinate system (lat/long vs projected)
# → Orders coordinates correctly for SaTScan

# From data frame
geo <- ss_geo(
  coords_df,
  loc_id = "id",
  coord1 = "longitude",
  coord2 = "latitude",
  coord_type = "latlong" # or "cartesian", or "auto"
)
```

Coordinate ordering is handled automatically:
- Lat/Long: SaTScan expects Latitude, Longitude (Y, X)
- Cartesian: SaTScan expects X, Y

### Parameter Management

### Parameter Management

Flexible configuration options:
```{r eval=FALSE}
# Method A: Object-Oriented (Recommended)
# - Auto-detects SaTScan version
# - Validates parameter names
# - Reusable across runs
my_opts <- prm_options(AnalysisType = 3, MonteCarloReps = 999)
satscanr(cas, geo, prm = my_opts)

# Method B: Quick Tweaks (Legacy/Interactive)
# - Good for single runs or quick checks
satscanr(cas, geo, AnalysisType = 3)

# Method C: Template Files (Replication)
# - Exact reproduction of GUI or previous runs
satscanr(cas, geo, prm_path = "space_time_template.prm")
```

Parameter validation:
```{r eval=FALSE}
# Validate against SaTScan's official templates
params <- list(
  AnalysisType = 3,
  ModelType = 1,
  MonteCarloReps = 999
)

prm_validate(params)
#> ✓ All parameters valid
#> ℹ Using template: Poisson space-time
```

## Available Input Types

| Function | File Type | Purpose |
|----------|-----------|---------|
| `ss_cas()` | `.cas` | Case counts by location/time |
| `ss_pop()` | `.pop` | Population denominators |
| `ss_geo()` | `.geo` | Spatial coordinates |
| `ss_ctl()` | `.ctl` | Controls (Bernoulli models) |
| `ss_grd()` | `.grd` | Custom scan centers |
| `ss_nwk()` | `.nwk` | Network edges (road networks) |
| `ss_nbr()` | `.nbr` | Neighbor rankings (wide format) |
| `ss_met()` | `.met` | Meta-locations (hierarchical grouping) |

## Documentation

- [Getting Started](vignettes/getting-started.html) - Complete workflow examples
- [Parameter Reference](vignettes/satscan-parameters.html) - All SaTScan parameters
- [Spatial Data Guide](vignettes/spatial-data.html) - Working with sf and coordinates
- [Report Issues](https://github.com/Songyosr/epidscan/issues)

## Development Status

This package is under active development (lifecycle: experimental). The API is stabilizing but may change in response to user feedback and testing.

## Citation
```r
citation("epidscan")

# Please also cite SaTScan:
# Kulldorff M. (1997). A spatial scan statistic. 
# Communications in Statistics: Theory and Methods, 26:1481-1496.
```

## License

MIT © Songyos Rajborirug
